---
title: "OPA for Wealth Tax"
author: ''
date: "1/28/2019"
output:
  html_document:
    code_folding: hide
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = "/Users/fhoces/Desktop/sandbox/opa-wealthtax")
library(tidyverse)
library(EnvStats)
library(here)
library(ggvis)
```


# Wealth tax analysis


```{r params, echo=TRUE, message=FALSE, warning=FALSE}
# - inputs: none
# - outputs: all the original source values
call_params_f <- function(){
    ################  
    ####### Research:
    ################  
    research_df <- read_csv("rawdata/research.csv")      #load data set that contains parameters from research
    # Elasticities
    ela1_so <- as.numeric(research_df[1,"param"])           # 0.5 - David. 2017
    ela2_so <- as.numeric(research_df[2,"param"])           # 0.5 - Jakobsen et al. 2018
    ela3_1_so <- as.numeric(research_df[3,"param"])         # 2   - Londono-Velez 2018
    ela3_2_so <-   as.numeric(research_df[4,"param"])       # 3   - Londono-Velez 2018
    ela4_1_so <- as.numeric(research_df[5,"param"])         # 23  - Brülhart et al. 2016
    ela4_2_so <- as.numeric(research_df[6,"param"])         # 34  - Brülhart et al. 2016
    
    ################
    ###### Data:
    ################ 
    # Tax base called from data (SCF and DINA) already accounts for the adjustments
    # due to population growth, tax avoidance, 
    # Number of tax payers SCF & DINA
    cum_numberTaxpayers_scf_so <- c(1100144, 298649, 72143, 34552, 7119,  2555,  671)
    cum_numberTaxpayers_dina_so <- c(673449, 194548, 78434, 32751, 10236, 4491, 1597) 
    # Tax base for brackets_po below
    cum_tax_base_scf_so <-  c(203.97, 121.63,  81.79, 59.18, 36.67, 27.46, 20.78)
    cum_tax_base_dina_so <- c(187.35, 135.38, 105.02, 80.11, 53.43, 36.94, 23.39)
    # Total wealth and number if households [SOURCE NEEDED]
    total_wealth_so <- 94e12   # [SOURCE]  
    total_hhlds_so <- 129.4e6  # [SOURCE]
    # Forbes: average across 400 and value for 400th billionare
    average_wealth_top_400_so <- 7.2e9
    wealth_last_400_so <- 2.1e9
    # Macro-economy/demographics
    inflation_so <- 0.025      # CBO/JCT
    population_gr_so <- 0.01   # CBO/JCT
    real_growth_so <- 0.02     # CBO/JCT
    
    ################ 
    #####  Guesswork:
    ################ 
    hhld_gr_so <- 0.009
    growth_wealth_so <- 0.055
    #As computed, the response to a 2% tax is a 16% tax elasticity. However the one
    #used in the report is 15%. Until we understand why, we will adjust by an explicit factor. 
    adj_factor1_so <- 0.01
    return( sapply( ls(pattern= "_so\\b"), function(x) get(x)) ) 
}
invisible( list2env(call_params_f(),.GlobalEnv) )


################ 
#####  Notes:
################ 
### Source ---------->  Input ---------->  Model ---------->  Policy Estimates (output)
###  (_so)              (_in)              (_mo)                (_pe)
### values            functions          functions              values
###                   & values           & values             
# - call_params_f - tax_elasticity_in_f  - tax_revenue_mo_f     - ten_year_revenue_pe
# - policy_f      - est_billionares_in_f - total_rev_mo_f       - ten_year_top_tax_pe
#                                        - ten_years_mo_f       - total_rev_pe
### arguments in functions should used "_var" and functions should "_f"

```


### 1 - Policy choices

The wealth tax applies to wealth above $50 million dollars, and follows the following structure:

```{r policy}
# - inputs: none
# - outputs: all the original source values
#### Policy:  
policy_f <- function(){
    # brackets_po
    brackets_po <- c(10, 25, 50, 100, 250, 500,  1000) * 1e6
    tax_rates_po <- c(  0,    0, 0.02,  0.02,  0.02,  0.02, 0.03) 
    starting_brack_po <- brackets_po[min(which(tax_rates_po>0))]
    next_increase_po <- brackets_po[min(which(tax_rates_po>0.02))]
    main_tax_po <- median(tax_rates_po)
    max_tax_po <- max(tax_rates_po)
    return( sapply( ls(pattern= "_po\\b"), function(x) get(x)) ) 
}
invisible( list2env(policy_f(),.GlobalEnv) )
knitr::kable(cbind("Bracket (millions of $)" = brackets_po/1e6,tax_rates_po) )  
```  


The tax begins at `r paste("$", starting_brack_po/1e6, sep = "")` million, and it taxes any additional wealth at `r paste(main_tax_po * 100, "%", sep = "")` rate. For those households with a wealth over `r paste("$", next_increase_po/1e9, sep = "")` billion, their additional wealth (above `r paste("$", next_increase_po/1e9, "b", sep = "")`) is taxed at a `r paste(max_tax_po * 100, "%", sep = "")` rate.
 
### 2 -  Compute tax elasticity

The tax-avoidance elasticity is computed as the average elasticity from four studies. 


Seim, David. 2017. “Behavioral Responses to an Annual Wealth Tax: Evidence from Sweden”, American Economic Journal: Economic Policy, 9(4), 395-421 and Jakobsen, Kristian, Katrine Jakobsen, Henrik Kleven and Gabriel Zucman. 2018. “Wealth Accumulation and Wealth Taxation: Theory and Evidence from Denmark” NBER working paper No. 24371, obtain small avoidance/evasion responses in the case of Sweden and Denmark in two countries with systematic third party reporting of wealth: a 1% wealth tax reduces reported wealth by less than 1%. Londono-Velez, Juliana and Javier Avila. “Can Wealth Taxation Work in Developing Countries? Quasi-Experimental Evidence from Colombia”, UC Berkeley working paper, 2018 show medium size avoidance/evasion responses in the case of Colombia where enforcement is not as strong: a 1% wealth tax reduces reported wealth by about `r paste(ela3_1_so, "-", ela3_2_so,"%", sep="")`. The study for Switzerland, Brülhart, Marius, Jonathan Gruber, Matthias Krapf, and Kurt Schmidheiny.“Taxing Wealth: Evidence from Switzerland,” NBER working paper No. 22376, 2016 is an outlier that finds very large responses to wealth taxation in Switzerland: a 1% wealth tax lowers reported wealth by `r paste(ela4_1_so, "-", ela4_2_so,"%", sep="")`. This extremely large estimate is extrapolated from very small variations in wealth tax rates over time and across Swiss cantons and hence is not as compellingly identified as the other estimates based on large variations in the wealth tax rate. Switzerland has no systematic third party reporting of assets which can also make tax evasion responses larger than in Scandinavia. 

```{r tax-elasticity, warning=FALSE, message=FALSE}  
# input: elasticity parameters from research, main tax, adjutment factor
# ouptut: final elasticity (final_ela_in), evasion parameter (evasion_param_in) 
tax_elasticity_in_f <- function(ela1_var = ela1_so, ela2_var = ela2_so, ela3_1_var = ela3_1_so, 
                                ela3_2_var = ela3_2_so, ela4_1_var = ela4_1_so, ela4_2_var = ela4_2_so,
                                main_tax_var = main_tax_po){
    final_ela_in <- mean(c(ela1_var, ela2_var, (ela3_1_var + ela3_2_var)/2, (ela4_1_var + ela4_2_var)/2))
    evasion_param_in <- main_tax_var * final_ela_in
    return(list("final_ela_in" = final_ela_in, 
                "evasion_param_in" = evasion_param_in))
}
invisible( list2env(tax_elasticity_in_f(),.GlobalEnv) ) 

#ls(pattern = "_in\\b|_in_")
```

The final `r paste(evasion_param_in * 100,"%", sep="")` tax avoidance/evasion response to a 2% wealth tax was computed as and average across these four studies. 

### 3 - Estimate number of billionares and their total tax base (over one billion)  
```{r est-billionares} 
# - inputs: evasion_param_in, wealth_last_400_so, average_wealth_top_400_so,  
# cum_numberTaxpayers_scf_so, cum_numberTaxpayers_dina_so, cum_tax_base_scf_so,  
# cum_tax_base_dina_so 
# - outputs: non_evaded_tax_in, totax_last_400_in, totax_average_400_in, totax_total_400_in, 
# pareto_scale_in, ratio1_in, missing_fraction_in, missing_taxbase_in, num_billionares_in, 
# final_taxbase_b_in, top_tax_base_in, cum_numberTaxpayers_scf_in, cum_numberTaxpayers_dina_in, 
# cum_tax_base_scf_in, cum_tax_base_dina_in 
est_billionares_in_f <- function(evasion_param_var = evasion_param_in, 
                                 wealth_last_400_var = wealth_last_400_so,
                                 average_wealth_top_400_var = average_wealth_top_400_so,
                                 cum_numberTaxpayers_scf_var = cum_numberTaxpayers_scf_so,
                                 cum_numberTaxpayers_dina_var = cum_numberTaxpayers_dina_so, 
                                 cum_tax_base_scf_var = cum_tax_base_scf_so, 
                                 cum_tax_base_dina_var = cum_tax_base_dina_so){
    non_evaded_tax_in <- (1 - evasion_param_var)
     
    # Post evasion taxable wealth
    totax_last_400_in <- non_evaded_tax_in * wealth_last_400_var/1e9                 
    totax_average_400_in <- non_evaded_tax_in * average_wealth_top_400_var/1e9       
    totax_total_400_in <- 400 * (totax_average_400_in - 1)                     # above one billion
      
    # Compute pareto scale of wealth distribution
    aux1 <- average_wealth_top_400_var/wealth_last_400_var
    pareto_scale_in <- aux1/(aux1-1) 
    ratio1_in <- round(aux1, 1)
    
    # Compute missing fraction of taxable wealth above 1 billion
    one_billion <- 1 
    missing_fraction_in <- ( totax_last_400_in / one_billion )^( pareto_scale_in - 1 ) - 1

    # Compute missing wealth (billions of dollars) to tax above one billion dollars
    missing_taxbase_in <- missing_fraction_in * totax_total_400_in

    # Compute total number of billionares. Extrapolation from forbes 400. 
    #Diff between original document (911) and below (906) is due to rounding in inputs 
    num_billionares_in <- 400 * ( totax_last_400_in / 1 ) ^ pareto_scale_in 
    final_taxbase_b_in <- totax_total_400_in + missing_taxbase_in

    #This is the taxed money (the aboves is the totale taxable) in billions  (minus one is just to match original results)
    top_tax_base_in <- final_taxbase_b_in * 0.01                                     # ALERT: Hard coded number
     
    #edit the number of billionares
    cum_numberTaxpayers_scf_in <- c(cum_numberTaxpayers_scf_var[1:6], num_billionares_in)
    cum_numberTaxpayers_dina_in <- c(cum_numberTaxpayers_dina_var[1:6], num_billionares_in)
    #edit the tax revenue for billionares    
    cum_tax_base_scf_in <- c(cum_tax_base_scf_var[1:6], top_tax_base_in)
    cum_tax_base_dina_in <- c(cum_tax_base_dina_var[1:6], top_tax_base_in)
    
    return( list( "non_evaded_tax_in" = non_evaded_tax_in, "totax_last_400_in" = totax_last_400_in, 
                  "totax_average_400_in" = totax_average_400_in, "totax_total_400_in" = 
                    totax_total_400_in, "pareto_scale_in" = pareto_scale_in, "ratio1_in" = ratio1_in, 
                  "missing_fraction_in" = missing_fraction_in, "missing_taxbase_in" = 
                    missing_taxbase_in, "num_billionares_in" = num_billionares_in, 
                  "final_taxbase_b_in" = final_taxbase_b_in, "top_tax_base_in" = top_tax_base_in, 
                  "cum_numberTaxpayers_scf_in" = cum_numberTaxpayers_scf_in, 
                  "cum_numberTaxpayers_dina_in" = cum_numberTaxpayers_dina_in, 
                  "cum_tax_base_scf_in" = cum_tax_base_scf_in, 
                  "cum_tax_base_dina_in" = cum_tax_base_dina_in ) )
}

invisible( list2env(est_billionares_in_f(),.GlobalEnv) )
#rm(list = c("aux1", "one_billion"))

```

For the billionaire surtax, the Forbes 400 reports that in 2018 the top 400 richest Americans had an average net worth of `r paste("$", average_wealth_top_400_so/1e9,sep="")` billion and that the 400th wealthiest had `r paste("$", wealth_last_400_so/1e9,sep="")` billion. Given the recent stock price declines, we assume that the wealth of the Forbes 400 in 2019 will be the same as in 2018. We also assume that their reported wealth for tax purposes would be `r paste(non_evaded_tax_in * 100,"%", sep="")` of the Forbes estimate due to avoidance and evasion. Hence the top 400 wealth taxpayers would report at least `r paste("$", round(totax_last_400_in, 1),sep="")` billion (= `r paste( non_evaded_tax_in," * ",wealth_last_400_so/1e9,sep="")`) and have an average reported wealth of `r paste("$", round(totax_average_400_in, 1),sep="")` billion (= `r paste( non_evaded_tax_in," * ",average_wealth_top_400_so/1e9,sep="")`). Hence, the Forbes 400 taxable base would be 400\*(6.1-1)=`r paste("$", totax_total_400_in, sep="")` billion or approximately `r paste("$", round(totax_total_400_in/1e3), sep="")` trillion.   

We use a standard Pareto interpolation technique, to estimate the billionaire tax base between \$1 billion and `r paste("$", round(totax_last_400_in, 1),sep="")` (below the Forbes 400).

We make the classical assumption that the tail of the wealth distribution is Pareto distributed. As the average wealth of the Forbes 400 (\$7.2b) is `r ratio1_in` times the threshold to belong to the Forbes 400 (\$2.1b), the corresponding Pareto parameter is a=3.4/(3.4-1)=`r round(pareto_scale_in,1)`. Standard calculations imply that the extra tax base between \$1bn and \$1.8bn (relative to the tax base above \$1.8bn) is [(1.8/1)^(a-1)-1] =`r round(missing_fraction_in, 3)` or approximately `r paste(round(missing_fraction_in*100), "%", sep="")` of the above `r paste("$", round(totax_total_400_in/1e3), sep="")` trillion. The number of families above \$1bn is estimated as 400 \* (1.8/1)^a=`r round(num_billionares_in)`, or about `r round(num_billionares_in/100) * 100`. The important point is that the vast majority of the billionaire surtax base (over three quarters) is in the Forbes 400, for which we have relatively good estimates of net worth.

Therefore, the billionaire surtax base is estimated at `r paste("$", round(final_taxbase_b_in/1e3, 2), sep="")` trillion (`r paste("$", round(totax_total_400_in/1e3),"t", sep="")` from those above `r paste("$", round(totax_last_400_in, 1),"b",sep="")` and `r paste("$", round(missing_taxbase_in/1e3,1),"t", sep="")` from those between 1  and `r paste("$", round(totax_total_400_in/1e3) ,"b",sep="")` from those above `r paste("$", round(totax_last_400_in, 1),sep="")`) and hence the billionaire surtax would raise `r paste("$", round(top_tax_base_in), sep="")` billion in 2019 from about `r round(num_billionares_in/100) * 100` billionaire families.


### 3.5 Compute tax Brackets: number of hhlds and taxable wealth [TO DO]

```{r tax-brackets, eval=FALSE, include = FALSE}

##### Reproducing do file
##### 
yr_val <- 2019
df_forbes <- read_dta("rawdata/materials/forbes_20112018_bdays.dta")
df_forbes1 <- df_forbes  %>% filter(forbes_yr == 2018) %>% 
  mutate("networth" = net_worthmillions * 1e6, 
           "weight" = 1, 
           "data" = "FB400") %>%
  select(data, networth, weight) %>% 
  filter( !is.na(networth) )
forbesmin <- min(df_forbes1$networth)
f400tot <- sum(df_forbes1$networth * df_forbes1$weight) / 1e12
cat("TOTAL FORBES NETWORTH 2018 (Tr) = ", f400tot,  "FORBES MIN WEALTH 2018 = ", forbesmin)

df_scf <- read_dta("rawdata/materials/rscfp2016.dta")
totw_scf <- sum(df_scf$networth * df_scf$wgt) / 1e12

cat("TOTAL SCF NETWORTH 2016  (Tr)", totw_scf)

df_scf <- df_scf %>% mutate("wgt" = round( wgt * 1.009^(yr_val-2016)))
totw <- sum(df_scf$networth * df_scf$wgt) / 1e12
totn <- sum(df_scf$wgt)


# Rescaling SCF to match total total wealth reported in DINA excluding the f400
df_scf1 <- df_scf %>% mutate("networth" = networth * ( 94 - f400tot ) / totw, 
                            "weight" = wgt, 
                            "data" = "SCF") %>%
  select(data, networth, weight)


# * DINA 2019 (complete  data  usdina$yr.dta built in usdina project, directory usdina/output/dinafiles)

# use hweal id dweght married using $root/SaezZucman2014/usdina/output/dinafiles/usdina$yr.dta, clear 
# save "$root/bookwebsite/wealthtaxsim/data/usdina$yr.dta"

df_dina <- read_dta("rawdata/materials/usdina2019.dta")
df_dina <- df_dina %>% 
  group_by(id) %>% 
  summarise("networth" = sum(hweal), 
            "weight" = mean(dweght) / 1e5 )

totw_dina <- sum(df_dina$networth * df_dina$weight) / 1e12
cat("TOTAL DINA NETWORTH 2019  (Tr) ", totw_dina)

totn_dina <- sum(df_dina$weight)

df_dina$data <- "DINA"

df_dina <- df_dina %>% 
  arrange(desc(networth)) 

#AQUI VOY
df_dina1 <- df_dina %>%
  mutate("aux_id" = floor((1:dim(df_dina)[1] + 3) / 5)) %>% 
  group_by(aux_id) %>% 
  summarise("weight" = sum(weight),
            "networth" = mean(networth)) %>% View() 
  mutate("data" = "DINA") %>% 
  select("data", "networth", "weight")

df <- rbind(df_forbes1, df_scf1, df_dina1)

df$weight <- with(df, ifelse(data=="SCF" | data=="DINA", 
                               round(weight/2), weight) )
df %>% 
  summarise( mean(weight), sd(weight) )

# replace weight=round(weight/2) if data=="SCF" | data=="DINA"
# drop if networth > `forbesmin' & (data=="SCF" | data=="DINA")
# sum networth [w=weight]
# display r(sum)*1e-12
# sum networth [w=weight] if networth>=1e+9
# display r(sum)*1e-12
# gsort -networth
# save $datawork/wealth.dta, replace



######



grid <- read.csv("taxBaseGrid.csv")  %>% filter(!is.na(gperc))
  # print(head(grid)) ## check that app has access to this file
  
  
  getTaxBasePerBracket=function(grid,brackets){
    ## brackets is lower end of each bracket
    brackets = c(brackets, 1e12) ## get last bracket
    #adding one dollar as the threshold represents the lower bound
    grid$group=cut(grid$thres + 1, brackets)
    toReturn = grid %>% group_by(group) %>% summarise(taxBase = sum(total)) %>% drop_na()
    return(toReturn)
  }
  
  getPeoplePerBracket=function(grid,brackets){
    brackets = c(brackets, 1e12) ## get last bracket
    grid$group=cut(grid$thres,brackets)
    toReturn = grid %>% group_by(group) %>% summarise(totalPeople=sum(nb)) %>% drop_na()
    return(toReturn)
  }
  
```



### 4 - Total tax revenue (Model)
```{r tax-revenue}  
# num_billionares_in <- round(400 * 1.8^1.4)
# top_tax_base <- 0.01 * ( 400 * ( (1-0.15) * 7.2 - 1 ) * 1.25 )
# Numbers in third bracket differ in 1 hhld 41636 instead of 4637
# numberTaxpayers <- c(640198, 171310, 41637, 24974, 5155, 2612, 911)
# 
    
# - inputs: cum_numberTaxpayers_dina_in, cum_numberTaxpayers_scf_in, num_billionares_in, cum_tax_base_dina_in, cum_tax_base_scf_in, brackets_po, starting_brack_po, main_tax_po
# - outputs: numberTaxpayers_mo, tax_base_mo, tax_base_dina_mo, tax_base_scf_mo, tax_base_total_mo, tax_rev_init_mo

#ADD COMMENTS
tax_revenue_mo_f <- function(cum_numberTaxpayers_dina_var = cum_numberTaxpayers_dina_in, cum_numberTaxpayers_scf_var = 
                               cum_numberTaxpayers_scf_in, num_billionares_var = num_billionares_in, 
                             cum_tax_base_dina_var = cum_tax_base_dina_in, cum_tax_base_scf_var =
                               cum_tax_base_scf_in, brackets_var = brackets_po, starting_brack_var = starting_brack_po,
                             main_tax_var = main_tax_po){
    aux1 <- (cum_numberTaxpayers_dina_var + cum_numberTaxpayers_scf_var)/2
    numberTaxpayers_mo <- c(-1 * diff(aux1, lag = 1), num_billionares_var)
    
    aux2 <- (cum_tax_base_dina_var + cum_tax_base_scf_var)/2
    tax_base_mo <- c(-1 * diff(aux2, lag = 1), top_tax_base_in) * 100
    
    tax_base_dina_mo <- cum_tax_base_dina_var[brackets_var == starting_brack_var]
    tax_base_scf_mo <- cum_tax_base_scf_var[brackets_var == starting_brack_var]
    
    target_hhlds_mo <- sum( numberTaxpayers_mo[brackets_var>=starting_brack_var] )
   
    tax_base_total_mo <- sum(tax_base_mo[brackets_var>=starting_brack_var])/1000
    tax_rev_init_mo <-   tax_base_total_mo * main_tax_var * 1000
    return(list("numberTaxpayers_mo" = numberTaxpayers_mo, "tax_base_mo" = tax_base_mo, "tax_base_dina_mo" =
                  tax_base_dina_mo,  "tax_base_scf_mo" = tax_base_scf_mo, "target_hhlds_mo" = 
                  target_hhlds_mo, "tax_base_total_mo" = tax_base_total_mo, "tax_rev_init_mo" = tax_rev_init_mo))
}
invisible( list2env(tax_revenue_mo_f(),.GlobalEnv) )


# Temporary values to be used in next paragraph             
target_hhlds_round <- format(round( sum( 
  numberTaxpayers_mo[brackets_po>=starting_brack_po] )/1000 
  ) * 1000, scientific = FALSE)
target_hhlds_round_scf <- format(round( sum( 
  cum_numberTaxpayers_scf_in[brackets_po==starting_brack_po] )/1000 
  ) * 1000, scientific=FALSE)
target_hhlds_round_dina <- format(round( sum( 
  cum_numberTaxpayers_dina_in[brackets_po==starting_brack_po] )/1000 
  ) * 1000, scientific=FALSE)
```


In 2019, there would be around `r  target_hhlds_round` households liable to the wealth tax (`r target_hhlds_round_scf` households according to the SCF data and `r  target_hhlds_round_dina` tax filers according to the DINA data). In both cases, this would be less than `r paste(round(target_hhlds_mo/total_hhlds_so * 100 , digits = 1), "%", sep = "")` of the `r round(total_hhlds_so/1e7)*10` million US households in 2019. The tax base above `r paste("$", starting_brack_po/1e6, sep="")` million would be `r paste("$", round(tax_base_scf_mo/10,1), sep="")` trillion based on the SCF data and `r paste("$", round(tax_base_dina_mo/10,1), sep="")` trillion based on the DINA data. Hence, the two data sources provide fairly close estimates, which we average as `r paste("$", round(tax_base_total_mo, digits = 1), sep="")` trillion, i.e. approximately 10% of total household net worth of \$94 trillion population-wide. A two percent tax on this base would raise `r paste("$", round(tax_rev_init_mo), sep="")` billion in 2019.

**First policy estimate**

```{r total-rev}
# - inputs: tax_rev_init_mo, top_tax_base_in
# - outputs: total_rev_pe
total_rev_mo_f <- function(tax_rev_init_var = tax_rev_init_mo, top_tax_base_var = top_tax_base_in) {
    total_rev_pe <- tax_rev_init_var + top_tax_base_var #PE
    return(list("total_rev_pe" = total_rev_pe))
}
invisible( list2env(total_rev_mo_f(),.GlobalEnv) )
```

The combination of the `r paste("%",100*main_tax_po, sep = "")` tax above `r paste("$",starting_brack_po/1e6, sep = "")` million and the billionaire surtax would raise `r paste(round(tax_rev_init_mo)," + ", round(top_tax_base_in), sep="")` = `r round(total_rev_pe)` billion in 2019.  

### 5 - Ten year projections  (talk about removing rounding)

```{r ten-years} 
# - inputs: inflation_so, population_gr_so, real_growth_so, total_rev_pe, top_tax_base_in
# - ouputs: discount_rate_mo, ten_year_factor_mo, ten_year_revenue_pe, ten_year_top_tax_pe
ten_years_mo_f <- function(inflation_var = inflation_so, population_gr_var = population_gr_so,
                           real_growth_var = real_growth_so, total_rev_var = total_rev_pe, 
                           top_tax_base_var = top_tax_base_in){
    discount_rate_mo <- inflation_var + population_gr_so + real_growth_so  
    ten_year_factor_mo <- sum( ( 1 + discount_rate_mo )^( 0:9 ) ) 

    ten_year_revenue_pe <- round(total_rev_pe) * ten_year_factor_mo     
    #ten_year_revenue_pe <- total_rev_pe * ten_year_factor_mo                      #PE
    ten_year_top_tax_pe <- top_tax_base_var * ten_year_factor_mo          
    #ten_year_top_tax_pe <- top_tax_base_var * ten_year_factor_mo                  #PE
    return( list("discount_rate_mo" = discount_rate_mo, "ten_year_factor_mo" = ten_year_factor_mo, 
           "ten_year_revenue_pe" = ten_year_revenue_pe, "ten_year_top_tax_pe" = ten_year_top_tax_pe) )
}

invisible( list2env(ten_years_mo_f(),.GlobalEnv) )

#TO DELETE
ten_year_factor_round <- round(ten_year_factor_mo) 



# test to run from the beginning (only functions)
if (FALSE) {
    rm(list = ls()[!(ls() %in% ls(pattern = "_f\\b"))])
    invisible( list2env(call_params_f(), .GlobalEnv) )
    invisible( list2env(policy_f(), .GlobalEnv) )
    invisible( list2env(tax_elasticity_in_f(), .GlobalEnv) )
    invisible( list2env(est_billionares_in_f(), .GlobalEnv) )
    invisible( list2env(tax_revenue_mo_f(), .GlobalEnv) )
    invisible( list2env(total_rev_mo_f(), .GlobalEnv) )
    invisible( list2env(ten_years_mo_f(),.GlobalEnv) )
    sapply(ls(pattern = "_pe\\b"), get)
}
```

To project tax revenues over a 10-year horizon, we assume that nominal taxable wealth would grow at the same pace as the economy, at `r paste(100 * discount_rate_mo,"%", sep="")` per year as in standard projections of the Congressional Budget Office or the Joint Committee on Taxation. This growth is decomposed into 2.5% price, 1% population growth, and 2% of real growth per capita. This implies that tax revenue over the 10 years 2019-2028 is `r ten_year_factor_round` times the revenue raised in 2019[^4]. This uniform growth assumption is conservative as the wealth of the rich has grown substantially faster than average in recent decades. The estimates by Saez and Zucman[^5] show that, from 1980 to 2016, real wealth of the top 0.1% has grown at 5.3% per year on average, which is 2.8 points above the average real wealth growth of 2.5% per year. Average real wealth of the Forbes 400 has grown even faster at 7% per year, 4.5 points above the average. The historical gap in growth rates of top wealth vs. average wealth is larger than the proposed wealth tax. Therefore, even with the wealth tax, it is most likely that top wealth would continue to grow at least as fast as the average.  

This 10-year projection implies that revenue raised by the progressive wealth tax would be `r paste(ten_year_factor_round, " * ", round(total_rev_pe), sep="")` = `r paste("$", round(ten_year_revenue_pe), sep = "")` billion, rounded to `r paste("$", round(ten_year_revenue_pe/1e3, 2), sep = "")` trillion. Out of these `r paste("$", round(ten_year_revenue_pe/1e3, 2), sep = "")` trillion, the billionaire surtax would raise `r paste(round(top_tax_base_in), " * ", ten_year_factor_round, sep="")` = `r paste("$", round(ten_year_top_tax_pe), sep ="")`, rounded to `r paste("$", round(ten_year_top_tax_pe/1e3,1), sep ="")` trillion.  

It is important to emphasize that our computations assume that the wealth tax base is comprehensive with no major asset classes exempt from wealth taxation. Introducing exemptions for specific asset classes would reduce the revenue estimates both mechanically and dynamically as wealthy individuals would shift their wealth into tax exempt assets. Because your proposal does not include any large exemptions, we do not believe our revenue estimate needs to be adjusted.

The figure below illustrates the distribution the tax burden across the population and, with detailed information on the fraction of households that will pay a wealth tax:

```{r figure, warning=FALSE}
wealth_aux <- 10:1000 * 1e6


numberTaxpayers <- c(640198, 171310, 41637, 24974, 5155, 2612, 911) ## eventually be able to change these based on other parameters
taxBase <- c(6716, 3510, 2376, 2460, 1285, 660, 2560) ## eventaully be able to change these based on other parameters
bracketNames <- c("$10m-$25m", "$25m-$50m", "$50m-$100m", "$100m-$250m", "250m-$500m", "$500m-$1bn", "1bn+")
#This function computes the total tax collected for a tax unit with wealth "wealth_var", applying "taxrates_var" to "brackets_var"
# - inputs: wealth, tax rates, brackets to tax
# - ouputs: total tax collected
getTaxRevenue <- function(wealth_var = wealth_aux, taxrates_var = tax_rates_po,
                          brackets_var = brackets_po) {
    ## expecting taxLevels in percentage
    # taxLevels <- taxLevels / 100
    if (length(brackets_var) != length(taxrates_var)){
      stop("Tax brackets and tax rates do not match")
    }
   # Compute max taxable wealth per bracket
    max_tax_per_brack <- c(diff(c(0, brackets_var)), 1e100)
   # Substract wealth minus tax bracket. If wealth above a given bracket (difference is larger than max taxable wealth), 
   # then assign max taxable wealth to that given bracket
    to_tax <- ifelse( wealth_var - c(0, brackets_var) > max_tax_per_brack, 
                      max_tax_per_brack, 
                      ( wealth_var - c(0,brackets_var) ) )   
   # If wealth if lower than a given bracket (difference between wealth and bracket is negative), then assign zero to that bracket  
    to_tax <- ifelse( to_tax<0, 0, to_tax )
    # Apply trax rates to each corresponding bracket and all together
    total_tax <- sum( to_tax * c(0, taxrates_var) )   
    return(total_tax)
}
  

xval <- seq(10e6, 1e9 + 1e8, by = 1e6)
    
#brackets_po <- c(0, 25, 50, 100, 250, 500, 1000) * 1e6
getGroup <- as.numeric(cut(xval, c(brackets_po, 1e12), include.lowest = TRUE))

toPlot <- cbind.data.frame(xval, getGroup)

toMatch <- cbind.data.frame(group = 1:7, tax = tax_rates_po)

toPlot2 <- merge(toPlot, toMatch, by.x = "getGroup", by.y = "group")

#lapply(x = 1:5, f(x,y), y = 6:10) = (1, 6:10); (2, 6:10);... ;(5, 6:10)  
toPlot2$averageInt <- sapply( toPlot2$xval, 
                              function(x) getTaxRevenue(wealth_var = x, 
                                                        taxrates_var = tax_rates_po, 
                                                        brackets_var = brackets_po) )

# Here is where the total tax payed by each individuals is transform into marginal tax rates
toPlot2$averageRate <- (toPlot2$averageInt / toPlot2$xval) * 100

toPlot2$id <- 1:nrow(toPlot2)
#browser()
#View(toPlot2)


#sapply( toPlot2$xval, function(x) getTaxRevenue(wealth_var = x, taxrates_var = tax_rates_po, brackets_var = brackets_po) )
 

##########
##########

#taxRate <- c(input$bracket1, input$bracket2, input$bracket3, input$bracket4, input$bracket5, input$bracket6, input$bracket7)
taxRate <- tax_rates_po *100


# These are mini data set that ggvis needs to create vertical lines
extra0 <- cbind.data.frame(x = rep(10e6, 2), y = c(0, taxRate[1]))
extra1 <- cbind.data.frame(x = rep(25e6, 2), y = c(0, taxRate[1]))
extra1b <- cbind.data.frame(x = rep(25e6, 2), y = c(0, taxRate[2]))
extra2 <- cbind.data.frame(x = rep(50e6, 2), y = c(0, taxRate[2]))
extra2b <- cbind.data.frame(x = rep(50e6, 2), y = c(0, taxRate[3]))

extra3 <- cbind.data.frame(x = rep(100e6, 2), y = c(0, taxRate[3]))
extra3b <- cbind.data.frame(x = rep(100e6, 2), y = c(0, taxRate[4]))

extra4 <- cbind.data.frame(x = rep(250e6, 2), y = c(0, taxRate[4]))
extra4b <- cbind.data.frame(x = rep(250e6, 2), y = c(0, taxRate[5]))
extra5 <- cbind.data.frame(x = rep(500e6, 2), y = c(0, taxRate[5]))
extra5b <- cbind.data.frame(x = rep(500e6, 2), y = c(0, taxRate[6]))
extra6 <- cbind.data.frame(x = rep(1e9, 2), y = c(0, taxRate[6]))
extra6b <- cbind.data.frame(x = rep(1e9, 2), y = c(0, taxRate[7]))

    
showMargin <- function(x) {
  # Walk through?
  # https://stackoverflow.com/questions/28396900/r-ggvis-html-function-failing-to-add-tooltip/28399656#28399656
  if (is.null(x)) return(NULL)
  data = toPlot2
  row <- data[data$id == x$id, ]
  paste0("Average Tax Rate: ", round(row$averageRate, 2), "%", sep = "")
}
    
    toPlot2$tax100 <- toPlot2$tax * 100
    
    toPlot2 %>%
      ggvis(x = ~xval, y = ~ tax100 ) %>%                                        # marginal tax
      layer_points() %>%
      layer_points(x = ~xval, y = ~averageRate, stroke := "red", key := ~id) %>%
      add_tooltip(showMargin, "hover") %>%
      layer_lines(x = ~xval, y = ~averageRate, stroke := "red") %>%
      layer_paths(data = extra1, ~x, ~y) %>%
      layer_paths(data = extra2, ~x, ~y) %>%
      layer_paths(data = extra3, ~x, ~y) %>%
      layer_paths(data = extra4, ~x, ~y) %>%
      layer_paths(data = extra5, ~x, ~y) %>%
      layer_paths(data = extra6, ~x, ~y) %>%
      layer_paths(data = extra0, ~x, ~y) %>%
      layer_paths(data = extra1b, ~x, ~y) %>%
      layer_paths(data = extra2b, ~x, ~y) %>%
      layer_paths(data = extra3b, ~x, ~y) %>%
      layer_paths(data = extra4b, ~x, ~y) %>%
      layer_paths(data = extra5b, ~x, ~y) %>%
      layer_paths(data = extra6b, ~x, ~y) %>%
      add_axis("x", title = "Wealth before taxes") %>%
      add_axis("y", title = "Tax rate (%)") %>%
      set_options(width = 1000, height = 500)

```



[Click here](https://fhoces.shinyapps.io/demo/) to try different policy proposals.   
[Click here/ADD URL WHEN ALLOWED](NULL) to see how the assumptions of the analysis affect the results.

[SAMPLE TEXT]**Note:** Our analysis complies with the highest levels of transparency and reproducibilty for open policy analysis proposed by the Berkeley Initiative for Transparency in Social Sciences. We invite contributors and critics of this analysis to follow similar standards. 

----  

# Original Letter to Sen. Warren  

Dear Senator Warren:   

We write regarding your proposal to impose a progressive annual wealth tax on American households with net worth (sum of all assets net of debts) above `r paste("$", starting_brack_po/1e6, "m", sep = "")` million. The tax would be 2% on the net worth above \$50 million with an additional 1% tax on net worth above \$1 billion. We estimate that about `r  format(round( sum( numberTaxpayers_mo[brackets_po>=starting_brack_po] )/1000 ) * 1000, scientific=FALSE)` American households (less than  `r paste(round(sum( numberTaxpayers_mo[brackets_po>=starting_brack_po] )/total_wealth_so * 100 , digits = 1), "%", sep = "")`) would be liable for the wealth tax and that the tax would raise around **`r paste("$", round(ten_year_revenue_pe/1e3, 2) - adj_factor1_so, sep = "")` trillion** over the ten-year budget window 2019-2028, of which `r paste("$", round(ten_year_top_tax_pe/1e3,1), sep ="")` trillion would come from the billionaire 1% surtax. The wealth tax would raise approximately 1.0% of GDP per year (\$210 billion relative to a \$21.1 trillion GDP in 2019).

## Details on the estimation

## Data Sources:

There are two main sources on the wealth of all American households: the Survey of Consumer Finances (SCF) from the Federal Reserve Board and the Distributional National Accounts (DINA) recently created by Piketty, Saez, and Zucman[^1], which estimates wealth by capitalizing investment income from income tax returns. The latest year available for each source is 2016. Both sources can be used to estimate the tax revenue of the 2% wealth tax above \$50 million. For the billionaire surtax, the best source is the Forbes 400 list of the richest 400 Americans. The latest year available is 2018.  

None of these sources provides perfect estimates but reassuringly, the SCF and DINA provide close estimates of the tax base above \$50 million. One of the important virtues of the progressive wealth tax is that it will generate much more accurate data to estimate and track the wealth of the wealthiest Americans.

## Methodology:

1 - We age the 2016 SCF and DINA microfiles to 2019 by inflating the number of households and wealth uniformly to match the aggregate projections for population and total household wealth from the Federal Reserve Board [SOURCE]. We also add the Forbes 400 to the SCF data. The total household net worth projection is \$94 trillion for 2019 (the SCF records a total household net worth of \$87 trillion in 2016).  

2 - Tax avoidance/evasion: recent research shows that the extent of wealth tax evasion/avoidance depends crucially on loopholes and enforcement. The proposed wealth tax has a comprehensive base with no loopholes and is well enforced through a combination of systematic third party reporting and audits. Therefore, the avoidance/evasion response is likely to be small. To be on the conservative side, we assume that households subject to the wealth tax are able to reduce their reported net worth by `r paste(evasion_param_in * 100,"%", sep="")` through a combination of tax evasion and tax avoidance. This is a large response in light of existing estimates[^2].

3 - In 2019, there would be around `r  format(round( sum( numberTaxpayers_mo[brackets_po>=50e6] )/1000 ) * 1000, scientific=FALSE)` households liable to the wealth tax (`r  format(round( sum( cum_numberTaxpayers_scf_in[brackets_po==50e6] )/1000 ) * 1000, scientific=FALSE)` households according to the SCF data and `r  format(round( sum( cum_numberTaxpayers_dina_in[brackets_po==50e6] )/1000 ) * 1000, scientific=FALSE)` tax filers according to the DINA data). In both cases, this would be less than `r paste(round(sum( numberTaxpayers_mo[brackets_po>=50e6] )/total_hhlds_so * 100 , digits = 1), "%", sep = "")` of the `r round(total_hhlds_so/1e7)*10` million US households in 2019. The tax base above \$50 million would be \$8.2 trillion based on the SCF data and \$10.5 trillion based on the DINA data. Hence, the two data sources provide fairly close estimates, which we average as `r paste("$", round(sum(tax_base_mo[brackets_po>=50e6])/1000, digits = 1  ), sep="")` trillion, i.e. approximately 10% of total household net worth of \$94 trillion population-wide. A two percent tax on this base would raise `r paste("$", round(sum(tax_base_mo[brackets_po>=50e6]) * 0.02), sep="")` billion in 2019.

4 - For the billionaire surtax, the Forbes 400 reports that in 2018 the top 400 richest Americans had an average net worth of `r paste("$", average_wealth_top_400_so/1e9,sep="")` billion and that the 400th wealthiest had `r paste("$", wealth_last_400_so/1e9,sep="")` billion. Given the recent stock price declines, we assume that the wealth of the Forbes 400 in 2019 will be the same as in 2018. We also assume that their reported wealth for tax purposes would be `r paste((1 - evasion_param_in) * 100,"%", sep="")` of the Forbes estimate due to avoidance and evasion. Hence the top 400 wealth taxpayers would report at least `r paste("$", round((1 - evasion_param_in) * wealth_last_400_so/1e9, 1),sep="")` billion (= `r paste( (1 - evasion_param_in)," * ",wealth_last_400_so/1e9,sep="")`) and have an average reported wealth of `r paste("$", round((1 - evasion_param_in) * average_wealth_top_400_so/1e9, 1),sep="")` billion (= `r paste( (1 - evasion_param_in)," * ",average_wealth_top_400_so/1e9,sep="")`). Hence, the Forbes 400 taxable base would be 400*(6.1-1)=\$2040 billion or approximately \$2 trillion. Using a standard Pareto interpolation technique, we estimate that the billionaire tax base between \$1 billion and \$1.8 billion (below the Forbes 400) would add 25% to the Forbes 400 tax base and would add about 500 families[^3]. Therefore, the billionaire surtax base is estimated at \$2.5 trillion and hence the billionaire surtax would raise \$25 billion in 2019 from about 900 billionaire families.

5 - The combination of the 2% tax above \$50 million and the billionaire surtax would raise \$187 + \$25 = \$212 billion in 2019.  

6 - To project tax revenues over a 10-year horizon, we assume that nominal taxable wealth would grow at the same pace as the economy, at 5.5% per year as in standard projections of the Congressional Budget Office or the Joint Committee on Taxation. This growth is decomposed into 2.5% price wealth_last_400_so, 1% population growth, and 2% of real growth per capita. This implies that tax revenue over the 10 years 2019-2028 is 13 times the revenue raised in 2019[^4]. This uniform growth assumption is conservative as the wealth of the rich has grown substantially faster than average in recent decades. The estimates by Saez and Zucman[^5] show that, from 1980 to 2016, real wealth of the top 0.1% has grown at 5.3% per year on average, which is 2.8 points above the average real wealth growth of 2.5% per year. Average real wealth of the Forbes 400 has grown even faster at 7% per year, 4.5 points above the average. The historical gap in growth rates of top wealth vs. average wealth is larger than the proposed wealth tax. Therefore, even with the wealth tax, it is most likely that top wealth would continue to grow at least as fast as the average.  

7 - This 10-year projection implies that revenue raised by the progressive wealth tax would be 13\*212=\$2756 billion, rounded to \$2.75 trillion. Out of these \$2.75 trillion, the billionaire surtax would raise \$25\*13=\$325 billion, rounded to \$0.3 trillion.  

8 - It is important to emphasize that our computations assume that the wealth tax base is comprehensive with no major asset classes exempt from wealth taxation. Introducing exemptions for specific asset classes would reduce the revenue estimates both mechanically and dynamically as wealthy individuals would shift their wealth into tax exempt assets. Because your proposal does not include any large exemptions, we do not believe our revenue estimate needs to be adjusted.

<!--  
Info from spreadsheet: 

Numbers from stata program wealthtax.do [to be reproduced]  

We start from SCF 2016 inflated to 2019 and adding Forbes 400 2018.   

We start from DINA tax data 2016 inflated to 2019.  

To inflate, we assume `r paste(hhld_gr_so*100,"%", sep="")` # household growth per year, wealth inflated uniformly to match 2019 aggregate total of $94Tr). Uniform discount of `r paste((final_ela_in * 0.02 - 0.01)*100,"%", sep="")` to account for tax evasion/avoidance when wealth tax in place. 
  
To do:   
 - remove all the rounded values.  
 - Incorporate wealthtax.do.    
 - replace any remaining hard coded values.    
 - 
  -->


[^1]: Piketty, Thomas, Emmanuel Saez, and Gabriel Zucman, “Distributional National Accounts: Methods and Estimates for the United States”, Quarterly Journal of Economics 133(2), 2018, 553-609. Data online at http://gabriel-zucman.eu/usdina/

[^2]: Seim, David. 2017. “Behavioral Responses to an Annual Wealth Tax: Evidence from Sweden”, American Economic Journal: Economic Policy, 9(4), 395-421 and Jakobsen, Kristian, Katrine Jakobsen, Henrik Kleven and Gabriel Zucman. 2018. “Wealth Accumulation and Wealth Taxation: Theory and Evidence from Denmark” NBER working paper No. 24371, obtain small avoidance/evasion responses in the case of Sweden and Denmark in two countries with systematic third party reporting of wealth: a 1% wealth tax reduces reported wealth by less than 1%. Londono-Velez, Juliana and Javier Avila. “Can Wealth Taxation Work in Developing Countries? Quasi-Experimental Evidence from Colombia”, UC Berkeley working paper, 2018 show medium size avoidance/evasion responses in the case of Colombia where enforcement is not as strong: a 1% wealth tax reduces reported wealth by about `r paste(ela3_1_so, "-", ela3_2_so,"%", sep="")`. The study for Switzerland, Brülhart, Marius, Jonathan Gruber, Matthias Krapf, and Kurt Schmidheiny.“Taxing Wealth: Evidence from Switzerland,” NBER working paper No. 22376, 2016 is an outlier that finds very large responses to wealth taxation in Switzerland: a 1% wealth tax lowers reported wealth by `r paste(ela4_1_so, "-", ela4_2_so,"%", sep="")`. This extremely large estimate is extrapolated from very small variations in wealth tax rates over time and across Swiss cantons and hence is not as compellingly identified as the other estimates based on large variations in the wealth tax rate. Switzerland has no systematic third party reporting of assets which can also make tax evasion responses larger than in Scandinavia.  Our `r paste(evasion_param_in * 100,"%", sep="")` tax avoidance/evasion response to a 2% wealth tax is based on the average across these four studies (2%*(.5+.5+2.5+28.5)/4=16%).

[^3]: We make the classical assumption that the tail of the wealth distribution is Pareto distributed. As the average wealth of the Forbes 400 (\$7.2b) is 3.4 times the threshold to belong to the Forbes 400 (\$2.1b), the corresponding Pareto parameter is a=3.4/(3.4-1)=1.4. Standard calculations imply that the extra tax base between \$1bn and \$1.8bn (relative to the tax base above \$1.8bn) is [(1.8/1)^(a-1)-1] =.27 or approximately 25%. The number of families above \$1bn is estimated as 400*(1.8/1)^a=911, or about 900. The important point is that the vast majority of the billionaire surtax base (over three quarters) is in the Forbes 400, for which we have relatively good estimates of net worth.

[^4]: With r=5.5%, we have [1+(1+r)+..+(1+r)^9]=[(1+r)^10-1]/r=`r round(sum((1 + 0.055)^(0:9)),1)`, approximately 13.

[^5]: Saez, Emmanuel and Gabriel Zucman, “Wealth Inequality in the United States since 1913: Evidence from Capitalized Income Tax Data”, Quarterly Journal of Economics 131(2), 2016, 519-578, updated series available at http://gabriel-zucman.eu/usdina/